---
title: "Untitled"
output: ioslides_presentation
date: "2024-03-25"
---




#### Understanding the Histogram of the Sampling Distribution of $\bar{X}$


--

The number of used cars sold in a summer week at Ally's Auto has a population distribution as shown in the probability histogram below. The population mean is also given.



```{r comment=NA, echo=FALSE, warning=FALSE, fig.align='center', fig.height=5}

sold <- 0:4
pct <- c(0.05,0.25,0.40,0.25,0.05)

barplot(pct, space = 0, xlab="\nNumber of used cars sold", ylab="Probability", main="Population")
axis(side=1,at = c(0.5,1.5,2.5,3.5,4.5), labels = 0:4)




```


---

What would the sampling distribution of the sample mean for a random sample of size *n = 3* summer weeks look like? *ALEKS: Use the slider to select the best answer.*


```{r comment=NA, eval=FALSE}

sample_3 <- replicate(10000, 
                      mean(
                        sample(0:4, 3, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_3, breaks = 10, xlim=c(0,4))

```


---

```{r comment=NA, echo=FALSE, fig.align='center'}

sample_3 <- replicate(10000, 
                      mean(
                        sample(0:4, 3, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_3, breaks = 10, xlim=c(0,4))

```


---

What would the sampling distribution of the sample mean for a random sample of size *n = 7* summer weeks look like? Use the slider to select the best answer.



```{r comment=NA, eval=FALSE}

sample_7 <- replicate(10000, 
                      mean(
                        sample(0:4, 7, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_7, breaks = 10, xlim=c(0,4))

```


---

```{r comment=NA, echo=FALSE, fig.align='center'}

sample_7 <- replicate(10000, 
                      mean(
                        sample(0:4, 7, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_7, breaks = 10, xlim=c(0,4))

```


---

What would the sampling distribution of the sample mean for a random sample of size *n = 32* summer weeks look like? Use the slider to select the best answer.



```{r comment=NA, eval=FALSE}

sample_32 <- replicate(10000, 
                      mean(
                        sample(0:4, 32, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_32, breaks = 10, xlim=c(0,4))

```


---

```{r comment=NA, echo=FALSE, fig.align='center'}

sample_32 <- replicate(10000, 
                      mean(
                        sample(0:4, 32, T, prob=c(0.05,0.25,0.40,0.25,0.05))
                        )
                      )

hist(sample_32, breaks = 10, xlim=c(0,4))

```


---

In summary:

--

- The sampling distribution of $\bar{X}$ is centered at the same value as $\mu$, regardless the sample size.

--

- $\mu_{\bar{x}} = \mu$

--

- The spread *(standard deviation)* of the sampling distribution of $\bar{X}$ decreases as the sample size increases.

--

- $\sigma_{\bar{x}} = \sigma/\sqrt{n}$

--

- The shape of the sampling distribution of $\bar{X}$ is Normal regardless the sample size **if** the population distribution is Normal and $\sigma$ is known.


---