---
title: "Introduction to Data Day 4"
date: "May 2, 2024"
output:
  xaringan::moon_reader:
    css: "myCSS.css"
    self_contained: true
    chakra: "remark-latest.min.js"
    nature:
      countIncrementalSlides: false
---


## Introduction to Data Day 4 

--

Analysis of TWO variables

--

For example, using the data frame **Orange** *(Run help(Orange))*, determine the answer to the following question: 

--

- Is trunk circumference related to the age of the tree? That is, do trees with large (or short) circumferences are much older? 

--

Step 1: Inspect the type of variables.

--

- What type of variables are *circumference* and *age*?

--

Step 2: Identify the appropriate type of analysis.

--

- Since *circumference* and *age* are quantitative, use scatterplot, correlation, and least-squares regression.


---

### Scatterplot

--

```{r comment=NA, eval=FALSE}

plot(age ~ circumference, data = Orange)

```

--

```{r comment=NA, echo=FALSE, fig.align='center', fig.height=6}

plot(age ~ circumference, data = Orange)

```

---

Add labels and improve visualization


--

```{r comment=NA, eval=FALSE}

plot(age ~ circumference, data = Orange, pch=19, xlab="Circunference (mm)",
     ylab="Age (Days)", main = "Grow of Orange Trees", col="blue")

```

--

```{r comment=NA, echo=FALSE, fig.align='center', fig.height=6}

plot(age ~ circumference, data = Orange, pch=19, xlab="Circunference (mm)", ylab="Age (Days)", main = "Grow of Orange Trees", col="blue")

```


---

### Assessment of the Linear Association

--

Compute the correlation coefficient (r) to measure the linear association between TWO quantitative variables.

--

```{r comment=NA, eval=FALSE}

cor(Orange$age, Orange$circumference)

```

--

```{r comment=NA, echo=FALSE}

cor(Orange$age, Orange$circumference)

```

--

- The correlation coefficient is positive and close to 1.


--

- There is a strong positive linear association between *circumference* and *age*.

--

- Old trees have the tendency to have large width.

--

A large correlation coefficient suggest the use of a linear model to describe the linear trend between *circumference* and *age*.

--

Before, it is important to identify which variable is the response (y) and predictor (x).


---

### Fitting the linear model

--

```{r comment=NA}

# Copy the linear model output
lm1 <- lm(age ~ circumference, data = Orange)

```

--

In this example, *circumference* is the predictor variable (x) and *age* the response variable (y). Why?

--

- If you see a tree, which one is easy to measure, the circumference or age?


--

```{r comment=NA, eval=FALSE}

# Display the linear model summary
summary(lm1)

```


---

```{r comment=NA, echo=FALSE}

# Display the linear model summary
summary(lm1)

```


--

- Line equation, age = 16.6 + (7.81)circumference

--

- Coefficient of determination $(R^2)$ is 0.8345


---

### Interpretation of Results

--

Line equation:

- Y-intercept = 16.6

--

- There is NO practical interpretation for the y-intercept 16.6.

--

- If there is not practical interpretation of the y-intercept, DO NOT mention it in your report.

--

- Slope = 7.81

--

- The orange trees get 7.81 days older ON AVERAGE for each millimeter increase in circumference.

--

Coefficient of determination $(R^2)$ = 0.8345

--

- The tree circumference can explain 83.45% of the variability of age. Therefore, it is strong predictor of the age of orange trees.

---

### Some Observations

--

- The linear regression analysis is not complete without the analysis of residuals.

--

- Therefore, the results previously reported are not conclusive. The analysis of residuals is a topic beyond the score of this course and data day.

--

- Comment in your report about the situations where these results can be helpful.

---

Using the data frame **chickwts** *(Run help(chickwts))*, determine the answer to the following question: 

--

- Do feed supplements  are effective on the growth of chickens? 

--

Step 1: Inspect the type of variables.

--

- What type of variables are *weight* and *feed*?

--

Step 2: Identify the appropriate type of analysis.

--

- Since *weight* and *feed* are quantitative and categorical respectively, use side-by-side boxplot, and summary of quantitative by categorical.

--

- Compare average of quantitative variable for each category.

--

- Identify unusual observations and similar spread across groups.


---

### Side-by-Side Box Plots

```{r comment=NA, fig.align='center', fig.height=6}

boxplot(weight ~ feed, data = chickwts)

```

---

Add title, appropriate label and other visual elements

--

```{r comment=NA, fig.align='center', fig.height=6}

boxplot(weight ~ feed, data = chickwts, xlab = "Type of Feed", 
        ylab = "Weight (lb)", main = "Chicken Weights by Feed Type", 
        col = "yellow")

```


---


```{r comment=NA}
# Summary by group type
tapply(chickwts$weight, chickwts$feed, summary)


```


---

Compare average (mean vs median)

--

- If no skewness is present, the mean and the median are about the same. Therefore compare means. Otherwise, compare medians.

--

- What group is greater (or smaller) on average?

--

- Why the averages are not the same? Were you expecting this pattern? 

--

Compare other characteristics:

--

- Is there any possible explanation for this similarity (or difference)

--

- What groups show evidence of unusual observations? Is there any possible explanation?


---
