---
title: "Hypothesis Testing for a Population Proportion"
date: "April 4, 2024"
output:
  xaringan::moon_reader:
    css: "myCSS.css"
    self_contained: true
    chakra: "remark-latest.min.js"
    nature:
      countIncrementalSlides: false
---

### Notation

--

- **p** is the population proportion of individuals who are in a specified category.

- **po** is the population proportion specified by Ho.

- **x** is the number of individuals in the sample who are in the specified category.

- **n** is the sample size.

- $\hat{p}$ is the sample proportion of individuals who are in the specified category. $\hat{p}$ = x/n.

--

### Reason for the assumptions

--

- The population must be much larger than the sample *(at least 20 times as large)*, so that the sampled items are independent.

--

- The assumption that both $np_o$ and $n(1 - p_o)$ are at least 10 ensures that the sampling distribution of $\hat{p}$ is approximately normal when we assume that Ho is true.


---

### Assumptions for Performing a Hypothesis Test for a Population Proportion.

--

We can perform a test whenever the sample proportion $\hat{p}$ is approximately normal. This will occur when the following assumptions are met:

--

- We have a simple random sample.

--

- The population is at least 20 times as large as the sample.

--

- The individuals in the population are divided into two categories.

--

- The values $np_o$ and $n(1 - p_o)$ are both at least 10.


---

### Performing a Hypothesis Test for a Population Proportion Using the P-Value Method.

--

- STEP 1: State the null and alternate hypotheses. The null hypothesis will have the form Ho: p = $p_o$. The alternate hypothesis, $H_1$, will be any of the following: p < $p_o$, p > $p_o$, p $\neq p_o$.

--

- STEP 2: If making a decision, choose a significance level $\alpha$.

--

- STEP 3: Compute the test statistic, Z = $\frac{\hat{p} - p_o}{\sqrt{\frac{p_o(1 - p_o)}{n}}}$.

--

- STEP 4: Compute the P-value. The P-value is the area under the standard normal curve. It depends on the type of test.

--

- STEP 5: Interpret the P-value. If making a decision, reject Ho if P-value $\leq \alpha$.

--

- STEP 6: State a conclusion.


---

### Testing Hypotheses for a Population Proportion Using the Critical Value Method.

--

- STEP 1: State the null and alternate hypotheses. The null hypothesis will have the form Ho: p = $p_o$. The alternate hypothesis will be any of the following: p < $p_o$, p > $p_o$, p $\neq p_o$.

--

- STEP 2: Choose a significance level $\alpha$ and find the critical value or values.

--

- STEP 3: Compute the test statistic, Z = $\frac{\hat{p} - p_o}{\sqrt{\frac{p_o(1 - p_o)}{n}}}$.

--

- STEP 4: Determine whether to reject Ho, as follows:

- - Left-tailed test: Reject if z $\leq -z_{\alpha}$.

- - Right-tailed test: Reject if z $\geq z_{\alpha}$.

- - Two-tailed test: Reject if z $\geq z_{\alpha}$ or z $\leq -z_{\alpha}$.

--

- STEP 5: State a conclusion.


---

The magazine Tech Worx reported that 92% of software engineers rate the company they work for as "a great place to work." As a veteran headhunter, you claim the percentage given in the report is not correct. In a survey of 210 randomly chosen software engineers, 188 rated the company they work for as "a great place to work."

Perform a hypothesis test to see if there is enough evidence, at the 0.10 level of significance, to support your claim that the proportion, p, of all software engineers who rate the company they work for as "a great place to work" is not 92%.


--


- Ho: p = 0.92

- $H_1:$ p $\neq$ 0.92

--

Checking assumptions:

--

- $np_o$ = (210)(0.92) = 193.2 > 10

--

- n(1 - $p_o$) = (210)(1 - 0.92) = 16.8 > 10


---

Compute the test statistic:

- Test statistic, Z = $\frac{\hat{p} - p_o}{\sqrt{\frac{p_o(1 - p_o)}{n}}}$

--

- Sample proportion, $\hat{p} = \frac{188}{210}$ = 0.895

--

- Z = $\frac{0.895 - 0.92}{\sqrt{\frac{0.92(1 - 0.92)}{210}}}$

--

- Z = -1.323

--

Using the P-value method for a two-tailed test.

--

- Use ALEKS calculator to find the left-tailed area: P(Z < -1.323)

--

- Because of symmetry, the right-tailed area is the same. Therefore, the P-value = 2*P(Z < -1.323) = 0.186 

--

- Since the P-value > $\alpha$, we fail to reject Ho.

---

Using the Critical Value method for a two-tailed test.

--

- Since $\alpha$ = 0.10, the critical values are determined for tails areas of 5% each.

--

- Using ALEKS calculator find the right-tailed critical value: $Z_{0.05}$.

--

- $Z_{0.05}$ = 1.645

--

- Because of symmetry, the left-tailed critical value is its negative, -1.645.

--

- Since the test statistic, z = -1.323, is not in the rejection region, we fail to reject Ho.


--

Conclusion

--

- There is not enough evidence to support the claim that the percentage of software engineers who rate the company they work for as "a great place to work" is not 92%.

---


At a nearby college, there is a school-sponsored website that matches people looking for roommates. According to the school's report, 43% of students will find a match their first time using the site. A writer for the school newspaper tests this claim by choosing a random sample of 160 students who visited the site looking for a roommate. Of the students surveyed, 54 said they found a match their first time using the site.


Complete the parts below to perform a hypothesis test to see if there is enough evidence, at the 0.05 level of significance, to reject the claim that the proportion, p, of all students who will find a match their first time using the site is 43%.

*a)*  State the null hypothesis (Ho) and the alternative hypothesis $(H_1)$ that you would use for the test.

--

- Ho: p = 0.43

--

- $H_1$: p $\neq$ 0.43


---

*b)* For your hypothesis test, you will use a Z-test. Find the values of np and n(1 - p) to confirm that a Z-test can be used. Here, n is the sample size and p is the population proportion that you are testing.

--

- n = 160, p = 0.43

--

- np = (160)(0.43) = 68.8

--

- n(1 - p) = (160)(1 - 0.43) = 91.2

--

- Since 68.8 and 91.2 are more than 10, the Z-test is appropriate.


--

*c)* Perform a Z-test and find the P-value.

--

- Z = $\frac{\hat{p} - p}{\sqrt{\frac{p(1 - p)}{n}}}$

--

- Z = $\frac{0.3375 - 0.43}{\sqrt{\frac{0.43(1 - 0.43)}{160}}}$

--

- Z = -2.3634


---

- Since this is a two-tailed test, we find the area below the Z-test -2.3634 and multiply by 2.

--

- P-value = 2*P(Z $\leq$ -2.3634)

--

- P-value = 2(0.009) = 0.018

--

*d)* Based on your answer to part (c), what can be concluded, at the 0.05 level of significance, about the claim made in the school's report?

--

- Since the P-value $\leq$ 0.05, reject Ho.

--

- There is enough evidence to reject the claim that 43% of students will find a match their first time using the site.

---

### Testing the Population Proportion in R Studio

--

Using the previous exercise as an example, the function **prop.test( )** will run the appropriate test.

--

```{r comment=NA, eval=FALSE}

prop.test(54, 160, 0.43, correct = F)

```

--

- The arguments in prop.test are the number of successes, sample size, null hypothesis value and the specification for the Z-test respectively. By default, **prop.test** is a two-tailed test with alpha 0.05.

--

```{r comment=NA, echo=FALSE}

prop.test(54, 160, 0.43, correct = F)

```

---

- We got the same P-value of 0.018 which leads to the rejection of the null hypothesis.

--

- The test statistics in this R-Studio output is X-squared = 5.5855. 
--

- The Z-test is $\sqrt{\text{X-squared}}$, so Z = $\sqrt{5.5855}$ = 2.3634.




