---
title: "Review for Exam 4"
date: "May 6, 2024"
output:
  xaringan::moon_reader:
    css: "myCSS.css"
    chakra: "remark-latest.min.js"
    self_contained: true
    nature:
      countIncrementalSlides: false
---




### The Chi-Square Distribution

--

```{r comment=NA, echo=FALSE, fig.width=12, fig.height=7, message=FALSE, warning=FALSE}

library(ggplot2)

x1 <- seq(0,40,0.01)
x2 <- rep(x1, 5)
y1 <- c(dchisq(x1,4), dchisq(x1,6), dchisq(x1,10), dchisq(x1,15), dchisq(x1,20))
z1 <- factor(rep(1:5,each=4001), labels=c(4, 6, 10, 15, 20))

df <- data.frame(x2,y1,z1)

ggplot(data=df, aes(x=x2, y=y1, color=z1)) +
  geom_path() +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7,0.8)) +
  labs(x=NULL, y=NULL, color="Degrees of \n Freedom")


```


---

Suppose that $\chi^2$ follows a chi-square distribution with 8 degrees of freedom. Use the ALEKS calculator to answer the following.

*a)* Compute $P(\chi^2 \leq 4)$. Round your answer to at least three decimals places.

--

- In ALEKS calculator, the function $P(\chi^2 > ... )$ finds the probability of above the number (right tail).

--

- Therefore, $P(\chi^2 \leq 4)$ = 1 - $P(\chi^2 > 4)$

--

- In ALEKS calculator, compute 1 - $P(\chi^2 > 4$ with df=8)

--

- Answer = 0.14287653952 $\approx$ 0.143

--

*b)* Find k such that $P(\chi^2 \geq k)$ = 0.1. Round your answer to at least two decimals places.

--

- In ALEKS calculator, use the function $\chi^2$_

--

- Type $\chi^2_{0.1; df=8}$

--

- Answer = 13.361566137 $\approx$ 13.36


---

Are birthdays "evenly distributed" throughout the year, or are they more common during some parts of the year than others? Owners of a children's toy store chain in the Northwest asked just this question. Some data collected by the chain are summarized in the table below.

The data were obtained from a random sample of 200 people. The birth date of each person was recorded, and each of these dates was placed into one of four categories: winter (December 21-March 20), spring (March 21-June 20), summer (June 21-September 20), and fall (September 21-December 20). The numbers in the first row of the table are the frequencies observed in the sample for these season categories. The numbers in the second row are the expected frequencies under the assumption that birthdays are equally likely during each season of the year. The bottom row of numbers gives the following value for each of the season categories.

$\frac{(f_o - f_e)^2}{f_e} = \frac{(Observed - Expected)^2}{Expected}$


---


**Part 1**

Fill in the missing values in the table. Round your responses for the expected frequencies to two or more decimal places. Round your $\frac{(f_o - f_e)^2}{f_e}$ responses to three or more decimal places.


. | Winter | Spring | Summer | Fall | Total
----|----|----|----|----|----
Observed, $f_o$ | 33 | 44 | 64 | 59 | 200
Expected, $f_e$ | - | 50 | 50 | - | -
$\frac{(f_o - f_e)^2}{f_e}$ | - | 0.720 | 3.920 | - | -

--

Are birthdays "evenly distributed" throughout the year, or are they more common during some parts of the year than others?

--

- Ho: $p_w = p_s = p_r = p_f = 0.25$

---

- By definition, $f_e$ = $(N)(p_o)$

--

- $f_e$ = (200)(0.25) = 50

--

- Winter: $\frac{(f_o - f_e)^2}{f_e}$ = $\frac{(33 - 50)^2}{50}$ = 5.78

--

- Fall: $\frac{(f_o - f_e)^2}{f_e}$ = $\frac{(59 - 50)^2}{50}$ = 1.62

--

. | Winter | Spring | Summer | Fall | Total
----|----|----|----|----|----
Observed, $f_o$ | 33 | 44 | 64 | 59 | 200
Expected, $f_e$ | 50 | 50 | 50 | 50 | -
$\frac{(f_o - f_e)^2}{f_e}$ | 5.78 | 0.720 | 3.920 | 1.62 | -


---

**Part 2**

Answer the following to summarize the test of the hypothesis that birthdays are equally likely during each season of the year. Use the 0.05 level of significance for the test.

*a)* Determine the type of test statistic to use.

--

- Chi-square. All expected frequencies at least 5.

--

- Degrees of freedom (k - 1) = 4 - 1 = 3

--

*b)* Find the value of the test statistic. (Round your answer to two or more decimal places.)

--

- $\sum \frac{(f_o - f_e)^2}{f_e}$ 

--

- 5.78 + 0.72 + 3.92 + 1.62 = **12.04**


---

*c)* Find the critical value. (Round your answer to two or more decimal places.)

--

- In ALEKS calculator, use the function $\chi^2$_

--

- Run $\chi^2_{0.05; df=3}$ 

--

- Answer = 7.8147279034 $\approx$ 7.81

--

*d)* Can we reject the hypothesis that birthdays are equally likely during each season of the year?

--

- Identify rejection region: [7.81, $\infty$)

--

- Since 12.04 $\in$ [7.81, $\infty$), reject Ho

---

The following table contains the first exam score and the second exam score for a random sample of 12 students in a large college class.

Exam 1 | 76.5 | 88.5 | 79.5 | 84.5 | 86.5 | 80.5 | 76 | 79 | 75 | 77.5 | 78 | 85.5
----|----|----|----|----|----|----|----|----|----|----|----
Exam 2 | 72.5 | 98 | 83.5 | 91.5 | 76.5 | 75.5 | 69 | 84.5 | 75 | 70.5 | 78 | 89.5
Difference | 4 | -9.5 | -4 | -7 | 10 | 5 | 7 | -5.5 | 0 | 7 | 0 | -4

At the 0.02 level of significance, can we conclude (by using the Wilcoxon signed-rank test) that there is a difference between the populations of first and second exam scores for students in this class?

*a)* Find the value of the test statistic.

--

- Ho: The population median of first and second exam scores are identical.

--

- $H_1$: The population median are NOT identical.


---


### The Signed-Rank Test

--

The Wilcoxon's signed-rank test is a **nonparametric** test designed to determine whether population **medians** differ when the data are in the form of **paired** samples. *(Ho: Same medians)*

--

It is therefore a nonparametric alternative to the t-test for matched pairs.

--

The main advantage of this nonparametric test is that it does not require that the population of differences have a specified distribution such as the normal distribution.

--

However, it does require that the distribution be approximately **symmetric**.

--

- After computing the difference, given in table, find their absolute values.

--

- |d| = 4, 9.5, 4, 7, 10, 5, 7, 5.5, 0, 7, 0, 4

--

- Rank the absolute values, but ignore any zero.

--

- Sorting: 4, 4, 4, 5, 5.5, 7, 7, 7, 9.5, 10; Ranking: 1, 2, 3, 4, 5, 6, 7, 8, 9, 10

---

Exam 1 | 76.5 | 88.5 | 79.5 | 84.5 | 86.5 | 80.5 | 76 | 79 | 75 | 77.5 | 78 | 85.5
----|----|----|----|----|----|----|----|----|----|----|----
Exam 2 | 72.5 | 98 | 83.5 | 91.5 | 76.5 | 75.5 | 69 | 84.5 | 75 | 70.5 | 78 | 89.5
Difference | 4 | -9.5 | -4 | -7 | 10 | 5 | 7 | -5.5 | 0 | 7 | 0 | -4
Absolute Value | 4 | 9.5 | 4 | 7 | 10 | 5 | 7 | 5.5 | 0 | 7 | 0 | 4
Rank | 2 | 9 | 2 | 7 | 10 | 4 | 7 | 5 | - | 7 | - | 2


--

Some considerations:

--

- Since the zeroes are ignored, ten numbers are ranked.

--

- The numbers 4 are rank in 1, 2, 3 places, but we take the rank "average", 2. Similarly, the number 7 is rank 7, average of ranks 6, 7, and 8.

--

Next step, assign the sign of the difference to the corresponding rank.

---

Exam 1 | 76.5 | 88.5 | 79.5 | 84.5 | 86.5 | 80.5 | 76 | 79 | 75 | 77.5 | 78 | 85.5
----|----|----|----|----|----|----|----|----|----|----|----
Exam 2 | 72.5 | 98 | 83.5 | 91.5 | 76.5 | 75.5 | 69 | 84.5 | 75 | 70.5 | 78 | 89.5
Difference | 4 | -9.5 | -4 | -7 | 10 | 5 | 7 | -5.5 | 0 | 7 | 0 | -4
Absolute Value | 4 | 9.5 | 4 | 7 | 10 | 5 | 7 | 5.5 | 0 | 7 | 0 | 4
Rank | 2 | 9 | 2 | 7 | 10 | 4 | 7 | 5 | - | 7 | - | 2
Signed rank | 2 | -9 | -2 | -7 | 10 | 4 | 7 | -5 | - | 7 | - | -2 

--

Add the absolute values of the negative signed ranks: 9 + 2 + 7 + 5 + 2 = 25

--

Add the positive ranks: 2 + 10 + 4 + 7 + 7 = 30

--

For a **two-tailed test**, the test statistic value, T, is the SMALLER of the two sums.

--

- Test statistic, **T = 25**.

---

*b)* Find the critical value.

--

- Use the Critical Values for the Wilcoxon Signed-Ranks Test table in ALEKS.

--

- The **n** is the number of nonzero differences, 10.

--

- Using the table for n = 10, $\alpha$ = 0.02 and two-tailed test, the critical value is **5**.

--

*c)*  Decide whether to reject Ho.

--

- Reject Ho when the test statistic is LESS THAN critical value.

--

- There is not enough evidence to suggest that the population exam scores are different. 


---


A psychologist wants to test whether there is any difference in puzzle-solving abilities between boys and girls. Independent samples of thirteen boys and nine girls were chosen at random. The boys took a mean of 36 minutes to solve a certain puzzle, with a standard deviation of 5.5 minutes. The girls took a mean of 37 minutes to solve the same puzzle, with a standard deviation of 5.3 minutes. Assume that the two populations of completion times are normally distributed, and that the population variances are equal. Can we conclude, at the 0.05 level of significance, that the mean puzzle-solving times for boys, $\mu_1$ differs from the mean puzzle-solving times for girls, $\mu_2$? $\phantom{ALEKS, OBJ7}$

--

*a)* State the null hypothesis, Ho and the alternative hypothesis, $H_1$.

--

- Setting the subindex 1 for boys and 2 for girls

--

- Ho: $\mu_1 - \mu_2$ = 0

--

- $H_1: \mu_1 - \mu_2 \neq$ 0

---

*b)* Determine the type of test statistic to use.

--

- T-test

--

- Degrees of freedom = 13 + 9 - 2 = 20

--

*c)* Find the value of the test statistic. 

--

- Since the population variances are assumed to be equal, we are going to pooled the standard deviation.

--

- t = $\frac{(\bar{X}_1 - \bar{X}_2) - (\mu_1 - \mu_2)}{S_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$

--

- $S_p$ is the pooled standard deviation: $\sqrt{\frac{(n_1 - 1)S_1^2 + (n_2 - 1)S_2^2}{(n_1 + n_2 - 2)}}$

--

- $S_p = \sqrt{\frac{(12)(5.5^2) + (8)(5.3^2)}{13 + 9 -2}}$ = 5.421

--

- t = $\frac{(36 - 37) - 0}{5.421\sqrt{1/13 + 1/9}}$ = -0.425


---


*d)* Find the p-value.

--

- Since this is a TWO-TAILED test, the p-value corresponds to the area below -0.425 and above 0.425.

--

- IN ALEKS *(area above 0.425)*: P(t > 0.425 with df=20)

--

- Right-tailed area is 0.3377

--

- By symmetry, the area below -0.425 is the same 0.3377

--

- The p-value is 0.3377 + 0.3377 = 0.6754

--

*e)* Can we conclude that the mean puzzle-solving times for boys differs from the mean puzzle-solving times for girls? Explain.

--

- Compare the p-value and alpha.

--

- Since alpha = 0.05, p-value > alpha.

--

- There is NOT enough evidence to conclude that the mean puzzle-solving times for boys differs from the mean puzzle-solving times for girls


---


---

A coin-operated coffee machine made by BIG Corporation was designed to discharge a mean of 7.1 ounces of coffee per cup. BIG has good reason to believe that the mean amount of coffee dispensed by the machine is greater than 7.1 ounces, and plans to do a statistical test of the claim that the machine is working as designed. BIG gathers a random sample of 75 amounts of coffee dispensed by the machine.

Suppose that the population of amounts of coffee dispensed by the machine has a standard deviation of 0.9 ounces and that BIG performs its hypothesis test using the 0.01 level of significance.

Based on this information, answer the questions below. Carry your intermediate computations to four or more decimal places, and round your responses as indicated.

*a)* What are the null hypothesis Ho and the alternative hypothesis $H_1$ that BIG should use for the test? 


--

- Ho: $\mu$ = 7.1

--

- $H_1$: $\mu$ > 7.1



---

*b)* What is the probability that BIG rejects the null hypothesis when, in fact, it is true? Round your response to two or more decimal places.

--

- By definition, this is $\alpha$ = Probability of Type I error.

--

- Since level of significance is 0.01, $\alpha$ = 0.01

--


*c)* Assuming that the actual value of $\mu$ is 7.43 ounces, what is the power of the test? Round your response to two or more decimal places. 

--

- By definition, Power = 1 - $\beta$

--

- $\beta$ is the probability of Type II error, fail to reject Ho when Ha is true.

--

- Step 1: Find the critical value corresponding to $\alpha$ = 0.01.

--

- In ALEKS calculator, $Z_{0.01}$ = 2.3263478773 $\approx$ = 2.3263

--

- Step 2: Find $\bar{X}$ = $\mu + (Z_{0.01})(\sigma/\sqrt{n})$

--

- $\bar{X}$ = $7.1 + (2.3263)(0.9\sqrt{75})$ = 7.3418


---

- Step 3: Find $\beta$ = $P(\bar{X}$ < 7.3418)

--

- $\beta$ = $P(\bar{X}$ < 7.3418) = $P(Z > \frac{7.3418 - 7.43}{0.9/\sqrt{75}})$ = P(Z < -0.8487)

--

- In ALEKS calculator, P(Z < -0.8487) = 0.20

--

- Step 4: Compute the Power of the test, 1 - $\beta$.

--

- Power = 1 - 0.20 = 0.80


--

*d)* Suppose that BIG decides to perform another statistical test using the same population, the same null and alternative hypotheses, and the same level of significance, but for this second test BIG chooses a random sample of size 125 instead of a random sample of size 75. Assuming that the actual value of $\mu$ is 7.43 ounces, how does the probability that BIG commits a Type Ⅱ error in this second test compare to the probability that BIG commits a Type Ⅱ error in the original test? 

--

- Start from Step 2 in part (c)

---

- Step 2: Find $\bar{X}$ = $\mu + (Z_{0.01})(\sigma/\sqrt{n})$

--

- $\bar{X}$ = $7.1 + (2.3263)(0.9\sqrt{125})$ = 7.2872

--

- Step 3: Find $\beta$ = $P(\bar{X}$ < 7.2872)

--

- $\beta$ = $P(\bar{X}$ < 7.2872) = $P(Z > \frac{7.2872 - 7.43}{0.9/\sqrt{125}})$ = P(Z < -1.7752)

--

- In ALEKS calculator, P(Z < -1.7752) = 0.037932434872 $\approx$ 0.04

--

- The probability of committing a Type Ⅱ error in the second test is less


---

Aragon Equipment, Inc. is a tractor manufacturer located in Ohio. It sells two types of tractors in each of four geographic regions. The contingency table below gives a summary of the sales data from this past year that Aragon executives are examining. A total of 500 Aragon tractors were sold during the year. The sales are classified according to two variables: geographic region of the sale ("Great Lakes", "Ohio Valley", "Mid-Atlantic", or "Corn Belt") and type of tractor ("The Incisor" or "The Caddy"). In the cells of the table are written the respective observed frequencies. In addition, three of the cells have blanks (A, B and C) beneath the observed frequencies. Fill in these blanks with the frequencies expected if the two variables, type of tractor and geographic region of the sale, are independent. Round your answers to two or more decimal places.

. | Great Lakes | Ohio Valley | Mid-Atlantic | Corn Belt | Total
----|----|----|----|----|----
The Incisor | 144 | 86 | 53 | 109 | 392
The Caddy | 48 | 16 | 21 | 23 | 108
Total | 192 | 102 | 74 | 132 | 500

---

. | Great Lakes | Ohio Valley | Mid-Atlantic | Corn Belt | Total
----|----|----|----|----|----
The Incisor | A | 86 | B | C | 392
The Caddy | 48 | 16 | 21 | 23 | 108
Total | 192 | 102 | 74 | 132 | 500

--

- Expected frequency for the ij-cell, $f_e = \frac{(Row_i)(Col_j)}{N}$

--

- A = $\frac{(392)(192)}{500}$ = 150.53

--

- B = $\frac{(392)(74)}{500}$ = 58.02

--

- C = $\frac{(132)(192)}{500}$ = 103.49


---

The human resources department of a large investment bank announced that the number of people it interviews monthly has a mean of 100 with a standard deviation of 16.3. The management of the bank suspects that the standard deviation exceeds 16.3. Suppose that the management wants to take a small sample of months and carry out a hypothesis test to see if its suspicions have support. State the null hypothesis Ho and the alternative hypothesis $H_1$ that it would use for this test. 

--

- Ho: $\sigma$ = 16.3

--

- $H_1$: $\sigma$ > 16.3

---


According to a high-profile realtor, houses in the sleepy town of Sun Beach have shown higher appreciation over the past three years than have houses in the bustling town of North Arden. To test the realtor's claim, an economist has found fourteen recently sold homes in Sun Beach and fourteen recently sold homes in North Arden that were owned for exactly three years. The following table gives the appreciation (expressed as a percentage increase) for each of the twenty-eight houses.

Sun Beach: 12.1, 7.8, 14.3, 8.0, 9.3, 9.9, 9.0, 13.9, 8.4, 12.1, 10.4, 11.9, 10.9, 13.3

North Arden: 9.3, 11.0, 9.1, 8.6, 6.4, 6.4, 10.1, 8.3, 10.8, 10.8, 10.2, 6.5, 9.3, 9.9

Assume that the two populations of appreciation rates are normally distributed and that the population variances are equal. Can we conclude, at the 0.05 level of significance, that the mean appreciation rate of houses in Sun Beach, $\mu_1$, is higher than $\mu_2$, the mean appreciation rate of houses in North Arden? 

*a)* State the null hypothesis Ho and the alternative hypothesis $H_1$

*b)* Determine the type of test statistic to use.

*c)* Find the value of the test statistic.

*d)* Find the p-value. 

*e)* Can we conclude that the mean appreciation rate of houses in Sun Beach, $\mu_1$, is higher than $\mu_2$, the mean appreciation rate of houses in North Arden?


---

### R Notes (Testing Mean difference in R)

--

```{r comment=NA, eval=FALSE}

# Store data
x <- c(12.1, 7.8, 14.3, 8.0, 9.3, 9.9, 9.0, 13.9, 8.4, 12.1, 10.4, 11.9, 10.9, 13.3)

y <- c(9.3, 11.0, 9.1, 8.6, 6.4, 6.4, 10.1, 8.3, 10.8, 10.8, 10.2, 6.5, 9.3, 9.9)

t.test(x, y, mu = 0, alternative = "greater", var.equal = T)


```


--

```{r comment=NA, echo=FALSE}

# Store data
x <- c(12.1, 7.8, 14.3, 8.0, 9.3, 9.9, 9.0, 13.9, 8.4, 12.1, 10.4, 11.9, 10.9, 13.3)

y <- c(9.3, 11.0, 9.1, 8.6, 6.4, 6.4, 10.1, 8.3, 10.8, 10.8, 10.2, 6.5, 9.3, 9.9)

t.test(x, y, mu = 0, alternative = "greater", var.equal = T)


```

--

- Since p-value > 0.05, reject Ho.

---





