---
title: "Hypothesis Test for a population Mean: Z test"
date: "March 28, 2024"
output:
  xaringan::moon_reader:
    css: "myCSS.css"
    self_contained: true
    chakra: "remark-latest.min.js"
    nature:
      countIncrementalSlides: false
---

### The Critical Value Method

--

- The critical value method is based on the idea that we should reject Ho if the value of the test statistic is "unusual" when we assume Ho to be true.

--

- In this method, we choose a **critical value**, which forms a "boundary" between values that are considered unusual and values that are not.

--

- The region that contains the unusual values is called the **critical region**. If the value of the test statistic is **IN** the critical region, we reject Ho.

--

- The critical value we choose depends on how small we believe a probability should be for an event to be considered unusual. In practice, an event with a probability of 0.05 or less is unusual.

--

- The probability that we use to determine whether an event is unusual is called the **significance level**.

--

- If we reject Ho after choosing a significance level $\alpha$, we say the result is **statistically significant** at the $\alpha$ level.


---


### Table of Critical Values *(for a Z test)*

Significance Level $\alpha$: | 0.10 | 0.05 | 0.02 | 0.01
----|----|----|----
Left-tailed | -1.282 | -1.645 | -2.054 | -2.326
Right-tailed | 1.282 | 1.645 | 2.054 | 2.326
Two-tailed | $\pm$ 1.645 | $\pm$ 1.96 | $\pm$ 2.326 | $\pm$ 2.576

--

A test is made of Ho: $\mu$ = 1 versus $H_1$: $\mu$ > 1. The value of the test statistic is Z = 1.85.

--

*a)* Is Ho rejected at the $\alpha$ = 0.05 level?

--

- The alternative hypothesis is $H_1$: $\mu$ > 1, so it is tight-tailed test.

--

- The critical value for $\alpha$=0.05 is $z_{\alpha}$=1.645. For the right-tailed test, we reject Ho if z $\geq$ $z_{\alpha}$. Because 1.85 > 1.645, we reject Ho at the $\alpha$ = 0.05 level.

--

*b)* If Ho rejected at the $\alpha$ = 0.01 level?


---

### The P-Value Method

--

Whereas the critical value method tells us only whether the test statistic was unusual or not, the P-value method tells us **exactly** how unusual the test statistic is. For this reason, the P-value method is the more often used in practice.

--

- The P-value is the **probability**, assuming that Ho is true, of observing a value for the test statistic that is as extreme as or more extreme than the value actually observed.

--

- The smaller the P-value, the stronger the evidence against Ho.

--

**Caution:** The P-value is NOT the probability that the null hypothesis is true.

--

To make a decision whether to reject Ho when using the P-value method:

--

- If P-value $\leq \alpha$, reject Ho. If P-value > $\alpha$, do not reject Ho.

--

- If P-value $\leq \alpha$, we say that Ho is rejected at the $\alpha$ level, or that the result is statistically significant at the $\alpha$ level.


---

Which provides stronger evidence against Ho: a P-value of 0.05 or a P-value of 0.50?

--

- A P-value of 0.05

--

A test is made of Ho: $\mu$ = 30 versus $H_1$: $\mu$ < 30. The test statistic is z = -1.28. Find and interpret the P-value.

--

- Using ALEKS calculator: P(Z $\leq$ -1.28)

--

- The answer is 0.1003


--

- If Ho is true, then the probability of observing a test statistic less than of equal to the value we actually observed *(-1.28)* is 0.1003.

--

- This result is not very unusual, so the evidence against Ho is not strong.

--

**Observation**: Since the alternative in this example is left-tailed, the ALEKS calculator function works well. However, you have to do 1 - P(Z $\leq$ _) if the alternative is right-tailed and 2
*P(Z $\leq$ -z) for the two-tailed test.


---

### Hypothesis Test for a population Mean: Z test

--

Assumptions (when $\sigma$ is known):

--

- We have a simple random sample.

--

- The sample size is large (n > 30), **OR** the population is approximately normal.

--

Once these assumptions are met, a hypothesis test can be performed using the following steps:

--

*1)* State the null and alternate hypothesis.

--

*2)* Choose a significance level $\alpha$, and find the critical value or values.

--

*3)* Compute the test statistic, Z = $\frac{\bar{x} - \mu_o}{\sigma/\sqrt{n}}$

--

*4)* Determine whether to reject Ho if the test statistic (Z) is in the critical region.


---


A company that manufactures batteries used in electric cars is reporting that their newest model of battery has a mean lifetime, $\mu$, of 18 years. To test the company's claim, a competitor has selected 64 of these batteries at random. The mean lifetime of the sample is 16.6 years. Suppose the population standard deviation of these lifetimes is known to be 4.7 years.

Is there enough evidence to reject the claim that the mean lifetime of the newest model is 18 years? Perform a hypothesis test, using the level of significance of 0.05.

--

- Ho: $\mu$ = 18

--

- $H_1$: $\mu \neq$ 18

--

- The population standard deviation, $\sigma$, is **KNOWN** and the sample size is large (generally n > 30) **OR** the population distribution is at least approximately normal.

--

- Z = $\frac{\bar{x} - \mu_0}{\frac{\sigma}{\sqrt{n}}}$

--

- Z = $\frac{16.6 - 18}{\frac{4.7}{\sqrt{64}}}$ = -2.383

---

The P-value approach

--

- In ALEKS calculator: 2*P(Z < -2.383)

--

- Because this is a two-tailed test, we multiply by two.

--

- P-value = 0.017

--

- Since the p-value is less than (or equal to) the level of significance, the null hypothesis is rejected. So, there is enough evidence to reject the claim that the mean lifetime of the newest model of battery is 18 years.

--

The rejection region approach *(Critical value method)*

--

- In ALEKS calculator: $Z_{0.025}$

--

- The output is 1.96 which is the Z critical value for the right tail. Since this is a two-tailed test, the other Z critical value is -1.96.

--

- Rejection region: $(\infty, -1.96)$ and $(1.96, \infty)$

--

- Since the test statistic, -2.383, is in the rejection region, reject Ho.

---


### Hypothesis Test for $\mu$ in R Studio

--

A magazine collects data each year on the price of a hamburger in a certain fast food restaurant in various countries around the world. The price of this hamburger for a sample of restaurants in Europe in January resulted in the following hamburger prices (after conversion to U.S. dollars)

5.19, 4.96, 4.05, 4.69, 5.28, 4.65, 4.13, 4.99, 5.16, 5.56, 5.34, 4.60

The mean price of this hamburger in the U.S. in January was $
4.63. For purposes of this exercise, assume it is reasonable to regard the sample as representative of these European restaurants. Does the sample provide convincing evidence that the mean January price of this hamburger in Europe is greater than the reported U.S. price? Test the relevant hypotheses using $\alpha$ = 0.05. 
 
--
 
 - Ho: $\mu$ = 4.63
 
 - $H_1$: $\mu$ > 4.63
 
---

Since the sample is small, the normal probability plot below suggest a t test.


```{r comment=NA, echo=FALSE, fig.align='center'}

x <- c(5.19, 4.96, 4.05, 4.69, 5.28, 4.65, 4.13, 4.99, 5.16, 5.56, 5.34, 4.60)
qqnorm(x, pch=19, col="red")
qqline(x)

```


---


- T = $\frac{\bar{X} - \mu_o}{S/\sqrt{n}}$, where t ~ T(df = n-1)

--

In this exercise: $\bar{X}$ = 4.883, S = 0.472, n = 12

--

- T = $\frac{4.883 - 4.63}{0.472/\sqrt{12}}$ = 1.86, where t ~ T(df = 11)

--

- Using the P-value approach, in ALEKS calculator: P(t > 1.86 with df = 11)

--

- P-value = 0.045. Since the p-value < $\alpha$, reject Ho.

--

```{r comment=NA, eval=FALSE}

# Copy the data
price <- c(5.19, 4.96, 4.05, 4.69, 5.28, 4.65, 4.13, 4.99, 5.16, 5.56, 
           5.34, 4.60)

# Run the hypothesis test
t.test(price, mu=4.63, alt = "greater")

```

---

```{r comment=NA, echo=FALSE}

# Copy the data
price <- c(5.19, 4.96, 4.05, 4.69, 5.28, 4.65, 4.13, 4.99, 5.16, 5.56, 
           5.34, 4.60)

# Run the hypothesis test
t.test(price, mu=4.63, alt = "greater")

```


---



