---
title: "Introduction to Data Day 1"
date: "February 6, 2025"
output:
  xaringan::moon_reader:
    css: "myCSS.css"
    chakra: "remark-latest.min.js"
    self_contained: true
    nature:
      countIncrementalSlides: false
---



### Numeric Summary of Quantitative Data

--

Quantitative data is best summarized with the five-number summary and the mean.

--

- The five-number summary are the minimum, first quartile, median, third quartile and maximum.

--

- The median is the number located in the middle of the sorted data. Therefore, half of the data is below the *median* and the other half above.

--

- The first quartile, *(Q1)*, is a number with one quarter (or 25%) of the data below and 75% above.

--

- Similarly, the third quartile, *(Q3)*, is a number with three quarter (or 75%) of the data below and 25% above.

---

In R, display the numeric summary of the variable **weight** in the data frame **chickwts**.

> Step 1: Open R

> Step 2: In R Console, type **library(Rcmdr)**, and hit ENTER.

Go to Step 5 if data is already opened

> Step 3: Open data, click on DATA > DATA IN PACKAGES > Read data set from an attached package

> Step 4: Enter name of data set: **chickwts**

> Step 5: Click on STATISTICS > SUMMARIES > NUMERICAL SUMMARIES

> Step 6: Select the variable **weight** and hit OK

--

<br>

```{r comment=NA, echo=F, message=F, warning=F}

library(Rcmdr)
numSummary(chickwts[,"weight", drop=F], statistics=c("mean","sd","IQR","quantiles"), quantiles=c(0,0.25,0.5,0.75,1))

```

---

Answer the following questions:

- What is the weight range in the sample?


- What is the approximate weight of the heavier chicken in the sample?


- What is the average weight?


- Are there chicken with unusual weight?


- Do the feed supplements help these chicken to growth fast?


---

### Graphical Summary of Quantitative Data

--

Plots of quantitative data help to visualize the distribution and unusual observations.

--

**HISTOGRAM**


- **Histograms** display the shape of the data distribution *(symmetry, skewness, modes)*.


- Histograms are similar to bar charts. Each bar is a **class intervals** with height equal to the interval's frequency. *(Frequency histogram)*

> In R, click on GRAPHS >  HISTOGRAM. Select variable. Choose the OPTION tab to add axes labels and title. Hit OK.

---

**BOX PLOT**


- A **Box plot** complements the histogram with the visualization of the five-number summary.


- Box plots are effective for the identification of **outliers** *(unusual observations)*

> In R, click on GRAPHS > BOXPLOT. Select variable. Choose the OPTION tab to add axes labels and title. Hit OK.



---


```{r comment=NA, fig.align='center', fig.width=7, fig.height=5, echo=F}

hist(chickwts$weight, 
        main="Frequency Histogram of Chicken Weight", 
        xlab="Weight (g)", ylab="Frequency")

```


---

```{r comment=NA, fig.align='center', fig.width=6, fig.height=5, echo=F}

boxplot(chickwts$weight, 
        main="Box plot of Chicken Weight", 
        xlab=NULL, ylab="Weight (g)")

```



---

### Introduction to Data Day 1

--

Step 1: Start R GUI

--

- Open R and type **library(Rcmdr)** in the R Console. Hit ENTER to lunch the R GUI.

--

Step 2: Prepare your MS Word document.

--

- In **Word** document, click on the INSERT > Header > Three Column Header

--

- On each column type: *your name*, *Section 1*, and your instructor name, *A. Falchettore*. 

--

- Save your Word document (FILE > SAVE AS) with the following format: your last name, your first initial, your student ID number, and the letters DD1. 

--

- So, if you are Sue Jones ID 99999999, your document would be saved as **joness99999999DD1**

--

Step 3: Access Brightspace and read the project instructions **BEFORE** downloading the data file for analysis


---

### Importing Data to R Studio (Practice)

--

There is a link to the **sleep** data set in BRIGHTSPACE. Download file **ONLY**. DO NOT OPEN.

- In R, click on DATA > IMPORT DATA > FROM EXCEL FILE. Write a name for the data set and hit OK.

- Browse your data and click OK.


---

### Instructions: 

--

Today, you are going to do two main things:

--

- Look at all the variables. Choose *ONE* quantitative and *ONE* categorical variables. Inspect them (*TWO* in total) using R. Describe each one as thoroughly as you can. *(Number of variables in actual Data Day 1 are different)*

--

- Tell me what question (for each of the chosen variables) you think was asked to result in the answers that were given.

--

The description for each variable should have this format:

- Name of Variable.

- Question that could have been asked to obtain the responses in this variable.

- Description of tables and charts in layman terms.
    
- Paste summary tables and charts from R Studio.

---

**Taking Screenshots of Tables and Charts**


- On the keyboard, press simultaneously the keys: **Windows Logo**, **Shift**, **S**.


- Select the area to take the screenshot.


- Paste in MS Word.


---


**Recommendations (Data Day)**


- Read the instructions CAREFULLY. Ask questions if something is not clear.


- Review all lecture notes and use them as reference.


- The key is to describe your results for somebody who does not understand statistics.


- Add title and labels to any chart in your report.


- Save your work OFTEN. Double check if your document was uploaded in BRIGHTSPACE correctly.


- Be organized and avoid any eventuality like slow computers, internet, or office 365.


